# Building a Database in R with RSQLite

In Chapter 6, we queried a database that already existed, someone else
had already built `frogs.sqlite` and populated it with data. This chapter
covers the other half of the picture: building a database from scratch,
starting from nothing but a set of CSV files, entirely from R.

This matters for your own final project. If your project ever grows
complex enough to warrant a relational database (see the "when not to
use a relational database" section in Chapter 5), you'll be the one
building it, and you'll almost certainly be starting from raw data files,
not an existing database.

We'll rebuild a small piece of the `frogs.sqlite` database ourselves,
from five CSV files: `frogs.csv`, `capture_sites.csv`, `tags.csv`,
`deployments.csv`, and `raw_gps_data.csv`. You can find these in the
`data/` folder of the course repository.

## Two functions, two purposes

So far, every query we've written has used `dbGetQuery()`, which
*retrieves* data and returns it as a data frame. Building a database
requires a second function, `dbExecute()`, for actions that *change* the
database itself, creating tables, inserting rows, deleting things. Both
functions take the same two arguments (a connection, and a SQL string),
they just serve different purposes: `dbGetQuery()` when you want data
back, `dbExecute()` when you want to make something happen.

## Creating a new, empty database

`dbConnect()` doesn't require a database to already exist. If the file
path you give it doesn't exist yet, it creates a brand new, empty
database there instead.

```{r, eval=FALSE}
library(DBI)

con <- dbConnect(RSQLite::SQLite(), "data/my_frogs.sqlite")
```

## Creating tables

We create each table with a `CREATE TABLE` statement, run through
`dbExecute()`, exactly the same SQL syntax you'd use directly in SQLite,
just wrapped in R.

```{r, eval=FALSE}
dbExecute(con, "CREATE TABLE frogs (
  frog_id INTEGER PRIMARY KEY,
  sex TEXT,
  first_capture_date TEXT,
  mortality_date TEXT
);")
```

## Loading a CSV and writing it to a table

Once a table exists, we read the corresponding CSV into R as an ordinary
data frame, then use `dbWriteTable()` to insert it. `dbWriteTable()`
takes the connection, the name of the table (as a string), and the data
frame to insert. Critically, we set `append = TRUE`, without it,
`dbWriteTable()` would overwrite the table entirely, along with the
`PRIMARY KEY` constraint we just defined.

```{r, eval=FALSE}
frogs <- read.csv("data/frogs.csv", stringsAsFactors = FALSE)
names(frogs)
```

Take a look at the column names before writing anything. This is a step
worth never skipping: `dbWriteTable()` matches CSV columns to table
columns by name, not by position, so if a name doesn't match exactly,
that column won't be recognized.

**Try it now:** Check whether `frogs.csv`'s column names match the
`frogs` table we just created.

```{r, eval=FALSE}
names(frogs)
```

You'll notice the first column is called `frog`, not `frog_id`, a
mismatch. Fix it before writing:

```{r, eval=FALSE}
names(frogs)[1] <- "frog_id"
dbWriteTable(con, "frogs", frogs, append = TRUE)
dbGetQuery(con, "SELECT * FROM frogs LIMIT 5;")
```

## Matching both names and order

Sometimes a CSV's columns are named correctly but in the wrong order,
which causes the same kind of mismatch. Let's create the
`capture_sites` table and look at its CSV.

```{r, eval=FALSE}
dbExecute(con, "CREATE TABLE capture_sites (
  site_id INTEGER PRIMARY KEY,
  site_name TEXT,
  latitude REAL,
  longitude REAL
);")

capture_sites <- read.csv("data/capture_sites.csv", stringsAsFactors = FALSE)
names(capture_sites)
```

**Try it now:** Compare that column order to the table definition above.
Both the names *and* the order need to line up before writing. Reorder
and rename as needed, then write:

```{r, eval=FALSE}
capture_sites <- capture_sites[, c("site", "site_name", "latitude", "longitude")]
names(capture_sites)[1] <- "site_id"
dbWriteTable(con, "capture_sites", capture_sites, append = TRUE)
dbGetQuery(con, "SELECT * FROM capture_sites;")
```

Repeat the same process for `tags`:

```{r, eval=FALSE}
dbExecute(con, "CREATE TABLE tags (
  tag_id INTEGER PRIMARY KEY,
  tag_model TEXT
);")

tags <- read.csv("data/tags.csv", stringsAsFactors = FALSE)
names(tags)
```

**Try it now:** `tags.csv` has columns named `tag` and `model`, neither
matches the table definition. Rename both, then write:

```{r, eval=FALSE}
names(tags) <- c("tag_id", "tag_model")
dbWriteTable(con, "tags", tags, append = TRUE)
dbGetQuery(con, "SELECT * FROM tags LIMIT 5;")
```

## Autoincrementing a primary key from R

Some tables don't have a natural, ready-made primary key in their CSV,
`deployments.csv`, for example, has no `deployment_id` column at all.
Rather than relying on SQL's `AUTOINCREMENT` keyword, we can just
generate a serial number ourselves, directly in R, before writing:

```{r, eval=FALSE}
dbExecute(con, "CREATE TABLE deployments (
  deployment_id INTEGER PRIMARY KEY,
  frog_id INTEGER,
  tag_id INTEGER,
  deployed_date TEXT,
  retrieved_date TEXT,
  FOREIGN KEY (frog_id) REFERENCES frogs(frog_id),
  FOREIGN KEY (tag_id) REFERENCES tags(tag_id)
);")

deployments <- read.csv("data/deployments.csv", stringsAsFactors = FALSE)
names(deployments)

# no deployment_id column exists yet -- create one
deployments$deployment_id <- 1:nrow(deployments)

# put it first, to match the table's column order
deployments <- deployments[, c("deployment_id", "frog_id", "tag_id",
                                "deployed_date", "retrieved_date")]

dbWriteTable(con, "deployments", deployments, append = TRUE)
dbGetQuery(con, "SELECT * FROM deployments LIMIT 5;")
```

**Try it now:** Load and write `raw_gps_data.csv` the same way, combining
everything so far: check the column names, rename and reorder as needed,
add a serial `reading_id`, then write.

```{r, eval=FALSE}
dbExecute(con, "CREATE TABLE raw_gps_data (
  reading_id INTEGER PRIMARY KEY,
  tag_id INTEGER,
  timestamp TEXT,
  raw_lat REAL,
  raw_lon REAL,
  FOREIGN KEY (tag_id) REFERENCES tags(tag_id)
);")

raw_gps_data <- read.csv("data/raw_gps_data.csv", stringsAsFactors = FALSE)
names(raw_gps_data)

raw_gps_data$reading_id <- 1:nrow(raw_gps_data)
raw_gps_data <- raw_gps_data[, c("reading_id", "tag", "timestamp", "lat", "lon")]
names(raw_gps_data) <- c("reading_id", "tag_id", "timestamp", "raw_lat", "raw_lon")

dbWriteTable(con, "raw_gps_data", raw_gps_data, append = TRUE)
dbGetQuery(con, "SELECT * FROM raw_gps_data LIMIT 5;")
```

## Deriving a new table from existing ones

This is where building a database in R really pays off. So far, every
table has come directly from a CSV. But `raw_gps_data` on its own isn't
very useful, it tells us which *tag* recorded each GPS fix, but not which
*frog* was wearing that tag at the time (a single tag can be reused
across multiple frogs and deployments over the course of a study).

We can derive a new table, `processed_gps_data`, by matching each raw GPS
reading to the deployment that was active at that exact moment, based on
matching `tag_id` and falling within that deployment's date range. This
uses `INSERT INTO ... SELECT`: instead of inserting values we type out
ourselves, we insert the *result of a query*.

```{r, eval=FALSE}
dbExecute(con, "CREATE TABLE processed_gps_data (
  location_id INTEGER PRIMARY KEY,
  frog_id INTEGER,
  timestamp TEXT,
  lat REAL,
  lon REAL,
  FOREIGN KEY (frog_id) REFERENCES frogs(frog_id)
);")

dbExecute(con, "
  INSERT INTO processed_gps_data (frog_id, timestamp, lat, lon)
  SELECT
    deployments.frog_id,
    raw_gps_data.timestamp,
    raw_gps_data.raw_lat,
    raw_gps_data.raw_lon
  FROM deployments
  LEFT JOIN raw_gps_data USING (tag_id)
  WHERE raw_gps_data.tag_id = deployments.tag_id
  AND (
    (
      (strftime(raw_gps_data.timestamp) >= strftime(deployments.deployed_date)) AND
      (strftime(raw_gps_data.timestamp) <= strftime(deployments.retrieved_date))
    )
    OR
    (
      (raw_gps_data.timestamp >= deployments.deployed_date) AND
      (deployments.retrieved_date IS NULL)
    )
  );
")

dbGetQuery(con, "SELECT * FROM processed_gps_data LIMIT 10;")
```

A few things worth unpacking in that `WHERE` clause:

- The first branch (inside the first pair of parentheses) handles the
  normal case: a deployment with both a start and an end date, where the
  GPS reading's timestamp needs to fall between the two.
- The second branch, after `OR`, handles a deployment whose
  `retrieved_date` is `NULL`, a tracker that was lost and never retrieved
  (recall Chapter 6's frog database, or your own bird database
  assignment, both have this same pattern). Without this second branch,
  every reading from a lost tracker's deployment window would be silently
  excluded, since comparing anything to `NULL` in SQL is never `TRUE`.

**Try it now:** Confirm every raw reading found a home. Since every
reading in this dataset happens to fall inside exactly one deployment
window, the two counts below should match:

```{r, eval=FALSE}
dbGetQuery(con, "SELECT COUNT(*) AS n FROM raw_gps_data;")
dbGetQuery(con, "SELECT COUNT(*) AS n FROM processed_gps_data;")
```

When you're done working with a database, close the connection:

```{r, eval=FALSE}
dbDisconnect(con)
```

## In-Class Activity: Building a Mini Database from Scratch

For this activity, you'll repeat the entire workflow from this chapter on
your own, from an empty database to a derived table, without following
along step by step.

### Setup

```{r, eval=FALSE}
library(DBI)

con <- dbConnect(RSQLite::SQLite(), "data/practice_frogs.sqlite")
```

### Your turn

1. **Create and populate the `frogs` table**, using `data/frogs.csv`.
   Remember to check the column names before writing.

<!--
ANSWER KEY (not visible in rendered output)

dbExecute(con, "CREATE TABLE frogs (
  frog_id INTEGER PRIMARY KEY,
  sex TEXT,
  first_capture_date TEXT,
  mortality_date TEXT
);")

frogs <- read.csv("data/frogs.csv", stringsAsFactors = FALSE)
names(frogs)
names(frogs)[1] <- "frog_id"
dbWriteTable(con, "frogs", frogs, append = TRUE)
-->

2. **Create and populate the `capture_sites` table**, using
   `data/capture_sites.csv`. Remember to check both names and column
   order.

<!--
ANSWER KEY (not visible in rendered output)

dbExecute(con, "CREATE TABLE capture_sites (
  site_id INTEGER PRIMARY KEY,
  site_name TEXT,
  latitude REAL,
  longitude REAL
);")

capture_sites <- read.csv("data/capture_sites.csv", stringsAsFactors = FALSE)
names(capture_sites)
capture_sites <- capture_sites[, c("site", "site_name", "latitude", "longitude")]
names(capture_sites)[1] <- "site_id"
dbWriteTable(con, "capture_sites", capture_sites, append = TRUE)
-->

3. **Create and populate the `tags` table**, using `data/tags.csv`.

<!--
ANSWER KEY (not visible in rendered output)

dbExecute(con, "CREATE TABLE tags (
  tag_id INTEGER PRIMARY KEY,
  tag_model TEXT
);")

tags <- read.csv("data/tags.csv", stringsAsFactors = FALSE)
names(tags)
names(tags) <- c("tag_id", "tag_model")
dbWriteTable(con, "tags", tags, append = TRUE)
-->

4. **Create and populate the `deployments` table**, using
   `data/deployments.csv`. This table has no ready-made primary key in
   the CSV, you'll need to generate one in R.

<!--
ANSWER KEY (not visible in rendered output)

dbExecute(con, "CREATE TABLE deployments (
  deployment_id INTEGER PRIMARY KEY,
  frog_id INTEGER,
  tag_id INTEGER,
  deployed_date TEXT,
  retrieved_date TEXT,
  FOREIGN KEY (frog_id) REFERENCES frogs(frog_id),
  FOREIGN KEY (tag_id) REFERENCES tags(tag_id)
);")

deployments <- read.csv("data/deployments.csv", stringsAsFactors = FALSE)
names(deployments)
deployments$deployment_id <- 1:nrow(deployments)
deployments <- deployments[, c("deployment_id", "frog_id", "tag_id",
                                "deployed_date", "retrieved_date")]
dbWriteTable(con, "deployments", deployments, append = TRUE)
-->

5. **Create and populate the `raw_gps_data` table**, using
   `data/raw_gps_data.csv`. Same as above: no ready-made primary key.

<!--
ANSWER KEY (not visible in rendered output)

dbExecute(con, "CREATE TABLE raw_gps_data (
  reading_id INTEGER PRIMARY KEY,
  tag_id INTEGER,
  timestamp TEXT,
  raw_lat REAL,
  raw_lon REAL,
  FOREIGN KEY (tag_id) REFERENCES tags(tag_id)
);")

raw_gps_data <- read.csv("data/raw_gps_data.csv", stringsAsFactors = FALSE)
names(raw_gps_data)
raw_gps_data$reading_id <- 1:nrow(raw_gps_data)
raw_gps_data <- raw_gps_data[, c("reading_id", "tag", "timestamp", "lat", "lon")]
names(raw_gps_data) <- c("reading_id", "tag_id", "timestamp", "raw_lat", "raw_lon")
dbWriteTable(con, "raw_gps_data", raw_gps_data, append = TRUE)
-->

6. **Derive `processed_gps_data`** from `deployments` and
   `raw_gps_data`, matching each raw reading to the frog whose deployment
   was active at that time. You're welcome to adapt the `INSERT INTO ...
   SELECT` statement from this chapter rather than writing it entirely
   from scratch, the point of this step is understanding *why* it works,
   not memorizing SQL syntax.

<!--
ANSWER KEY (not visible in rendered output)

dbExecute(con, "CREATE TABLE processed_gps_data (
  location_id INTEGER PRIMARY KEY,
  frog_id INTEGER,
  timestamp TEXT,
  lat REAL,
  lon REAL,
  FOREIGN KEY (frog_id) REFERENCES frogs(frog_id)
);")

dbExecute(con, "
  INSERT INTO processed_gps_data (frog_id, timestamp, lat, lon)
  SELECT
    deployments.frog_id,
    raw_gps_data.timestamp,
    raw_gps_data.raw_lat,
    raw_gps_data.raw_lon
  FROM deployments
  LEFT JOIN raw_gps_data USING (tag_id)
  WHERE raw_gps_data.tag_id = deployments.tag_id
  AND (
    (
      (strftime(raw_gps_data.timestamp) >= strftime(deployments.deployed_date)) AND
      (strftime(raw_gps_data.timestamp) <= strftime(deployments.retrieved_date))
    )
    OR
    (
      (raw_gps_data.timestamp >= deployments.deployed_date) AND
      (deployments.retrieved_date IS NULL)
    )
  );
")
-->

7. **Confirm your work**: query `processed_gps_data` and check that the
   row count matches `raw_gps_data`.

<!--
ANSWER KEY (not visible in rendered output)

dbGetQuery(con, "SELECT COUNT(*) AS n FROM raw_gps_data;")
dbGetQuery(con, "SELECT COUNT(*) AS n FROM processed_gps_data;")

Both should return 160.
-->

### What to submit

Submit an RMarkdown document (or plain R script) containing every
`CREATE TABLE`, `dbWriteTable()`, and the final `INSERT INTO ... SELECT`
statement you used, along with the output confirming your row counts
match.

## References

- `DBI` package documentation: <https://dbi.r-dbi.org/>
- `RSQLite` package documentation: <https://rsqlite.r-dbi.org/>
