# Assignments

This chapter collects each week's assignment in one place for easy
reference. New assignments will be added here as the semester progresses.

## Week 1: Software Setup and Final Project Topic Selection

Before our first class session, please install R and RStudio and verify
your installation.

## Software Requirements and Installation Instructions {-}

Before our first class session, please install the following software.
Come to class with everything installed and verified — we'll do a quick
check together, but coming prepared means we can dive into the material
right away.

### R {-}

R is a free software environment for statistical computing and graphics.
Installing or updating R is a separate process from installing or updating
RStudio (below) — you need both.

- **Windows:** Download the latest version at <https://cran.r-project.org/bin/windows/base/>
- **Mac OS:** Download the latest version at <https://cran.r-project.org/bin/macosx/>
- **Linux:** Follow the instructions for your distribution at <https://cloud.r-project.org/bin/linux/>

### RStudio {-}

RStudio is a free Integrated Development Environment (IDE) for R. Installing
or updating RStudio is separate from installing or updating R.

Download it from <https://posit.co/download/rstudio-desktop/>.

## The RStudio Interface {-}

When you open RStudio, you'll see the window divided into four panes. Each
one serves a different purpose:

**Source (top left)**
This is where you write and edit scripts, RMarkdown documents, and other
files. Code you write here isn't run automatically — you need to send it to
the Console (below) to execute it. Keeping your code in the Source pane,
rather than typing directly into the Console, is what makes your work
reproducible: it's saved, shareable, and can be re-run from start to finish.

**Console (bottom left)**
This is where R actually executes code and shows you the results. You can
type commands directly into the Console, but anything you type there isn't
saved anywhere — so it's best used for quick, one-off checks rather than your
actual analysis.

**Environment / History (top right)**
The Environment tab shows you all the objects (data frames, variables,
functions, etc.) currently loaded in your R session, along with their type
and a preview of their contents. The History tab keeps a running log of every
command you've executed in the Console during your session.

**Files / Plots / Packages / Help (bottom right)**
This pane is multi-purpose. *Files* lets you browse your project's folder
structure. *Plots* displays any graphs or visualizations you create. *Packages*
shows which R packages are installed and lets you load or update them. *Help*
displays documentation when you look up a function.

You can resize any pane by dragging its borders, and rearrange their
positions under **Tools > Global Options > Pane Layout** if you'd prefer a
different arrangement.

### Required R Packages {-}

We'll be using the following R packages throughout the course: `tidyverse`,
`rmarkdown`, `bookdown`, and `renv`. Install them, along with their dependencies, by running the
following in the R console:

```r
install.packages(c("tidyverse",
                    "rmarkdown",
                    "bookdown",
                    "renv"), dependencies = TRUE)
```

### Verifying Your Installation {-}

To confirm everything is working, open RStudio and run the following in
the console:

```r
R.version.string
library(tidyverse)
library(bookdown)
library(renv)
```

If R prints a version number and the `tidyverse` loads without error
messages, you're all set.

### Dataset eligibility criteria

Your dataset must meet the following criteria:

- **Public availability:** The dataset must be publicly accessible — for
  example, from a government agency, public data repository, or a published
  study's supplementary materials. No proprietary, restricted-access, or
  personally identifiable data.
- **Minimum size:** At least **100 rows (observations)** and **10 variables
  (columns)**.
- **Rationale:** A short written explanation (3–5 sentences) of why this
  dataset interests you and what you hope to learn from it.
- **Exploratory questions:** At least 2–3 questions you want to answer using
  this dataset by the end of the course.


### What to submit

Go to Canvas, and submit a screen shot of your R console output for the 
"Week 1" assignment that shows your successful download.

Then in the text box, submit the following:

- A brief description of your dataset, including an overview of how the data were collected, how many observations there are, and how many variables were measured.
- A short written explanation (3–5 sentences) of why this dataset interests you and what you hope to learn from it.
- At least 2–3 questions you want to answer using this dataset by the end of the course.
- A brief description on any data cleaning you anticipate needing to do, based on an initial look at the file.

## Week 2: Project Directory Structure and Git Setup

Now that you've chosen your final project dataset, it's time to set up the
project itself as a properly structured, version-controlled repository.

### What to do

**1. Design your directory structure.**

Based on what we covered in Chapter 1, plan out a folder structure for your
final project. At minimum, your structure should separate raw data,
scripts, and output — but feel free to adapt it to fit your specific
project.

**2. Create the folders locally.**

Make a new folder for your project (name it something descriptive, e.g.
`final-project`), then create your subfolders inside it. You can do this in
Finder/File Explorer, or from the terminal. For example, from inside your
project folder:

```
mkdir data
mkdir data/raw
mkdir data/processed
mkdir scripts
mkdir output
mkdir output/figures
```

(Adjust folder names to match your own planned structure — this is just an
example.)

**3. Turn it into a Git repository.**

Navigate into your project folder in the terminal (recall the drag-and-drop
or Option+right-click tricks from Chapter 2 if you need to find the path),
then run:

```
git init
```

**4. Add a `.gitignore` file.**

Create a new plain-text file named exactly `.gitignore` in your project's
top-level folder (see Chapter 2 for instructions specific to your
operating system). At minimum, include the following:

```
.Rhistory
.Rproj.user/
.DS_Store
```

Add any additional rules relevant to your own project — for example, if
you'll be working with large raw data files you don't want tracked.

**5. Add a README.**

Create a new plain-text file named `README.md` in your project's top-level
folder. At minimum, it should include:

- Your project title and a one-paragraph description of what it's about
- A brief explanation of your folder structure (what lives where)
- A note on your data source (a link or citation)

**6. Make your first commit.**

From inside your project folder, run:

```
git add .
git commit -m "Initial project structure"
```

### What to submit

Submit a short document (PDF or Word) containing:

1. A screenshot showing your folder structure.
2. A brief paragraph (3–5 sentences) explaining your folder structure —
   what each top-level folder is for and why you organized it this way.

## Week 3: Creating a GitHub Repository for Your Project

In Week 2, you created a local Git repository for your final project. This
week, you'll put that repository on GitHub so it's backed up remotely and
ready for collaboration.

### What to do

**1. Create a new repository on GitHub.**

- Go to <https://github.com> and log in.
- Click the **+** icon in the top-right corner and select **New repository**.
- Fill out the repository details:
  - **Repository name:** Use a short, descriptive name for your project
    (e.g., `my-lastname-final-project`)
  - **Description:** One sentence describing your project topic
  - **Visibility:** Choose **Public** (unless your data requires restricted
    access, in which case choose **Private**)
  - **Do NOT** check "Add a README file," "Add .gitignore," or "Choose a
    license" — your local repo already has these files, and we'll connect
    it directly
- Click **Create repository**. GitHub will show you a page with setup
  instructions — keep this page open.

**2. Connect your local repo to GitHub.**

Copy the repository URL from the GitHub page (it will look like
`https://github.com/yourusername/your-repo-name.git`). Then, from inside
your project folder, run the following, replacing `<repo-url>` with the URL
you copied:

```
git remote add origin <repo-url>
git branch -M main
git push -u origin main
```

If prompted for a username and password, enter your GitHub username. For
the password, use a Personal Access Token (PAT) instead of your account
password (see Chapter 3 for a refresher on generating one).

**3. Verify it worked.**

Refresh your repository's page on GitHub and confirm that your project
files — including your `.gitignore` and `README.md` — appear there.

### What to submit

Submit the following:

1. A screenshot of your repository's GitHub page showing your files.
2. The URL to your GitHub repository.

## Week 4: Designing a Tidy Spreadsheet for a Bear Research Case Study

This week, you'll apply the tidy data principles from Chapter 4 to a new
scenario — not your own final project dataset, but a case study describing
a black bear research project. Real field data collection is often
described informally, in field notes or verbal summaries, rather than
handed to you as a ready-made spreadsheet. Your job is to translate a
messy, informal description of a research project into a properly
structured, tidy spreadsheet design.

### The case study

A research team is studying a black bear population in a mountainous study
area. Here's how they describe their data collection process:

> "We trap bears using culvert traps and snares placed at a handful of
> fixed trap sites scattered across the study area. Every time we catch a
> bear — whether it's the first time or a recapture — we record the date,
> which trap site it was caught at, and the bear's ID (new bears get a new
> ID; if we recognize an ear tag from a previous capture, we reuse that
> ID). At first capture, we record the bear's sex. Every time we handle a
> bear, whether it's a new capture or a recapture, we take its weight and
> body length, and we note its age class (cub, yearling, or adult). Over
> the course of the study, we catch dozens of different bears, some of
> them multiple times, at several different trap sites, across several
> field seasons."

### What to do

**1. Identify the variables and observations.**

Read through the case study carefully. For each piece of information being
collected, ask: what is being measured (a variable)? Then think about what
a single row in your table should represent — one bear? One capture event?
Be careful here: a bear that's caught three times shouldn't be squeezed
into a single row, since its weight, length, age class, trap site, and
date are all different at each capture.

**2. Design your table.**

Sketch out the column headers your table needs, keeping in mind that the
same bear ID will appear on more than one row (once per capture), while
other information (like sex) stays the same across all of that bear's
rows. This is expected and fine — it's how a single tidy table
accommodates repeated measurements on the same individual.

**3. Build the actual spreadsheet template.**

In Excel or Google Sheets, create your table with only column headers
filled in — you don't need to invent or enter any actual data rows. Column
headers should be computer-readable (following the naming conventions from
Chapter 1: no spaces, no special characters, consistent capitalization).

Pay close attention to the tidy data pitfalls from Chapter 4 as you design
your headers: don't cram two variables into one column (like age class and
sex), don't let a single cell need to hold more than one value, and make
sure your table's unit of observation is consistent row to row.

**4. Write a short justification.**

In a separate document (or as a second tab/sheet), briefly explain:

- What a single row in your table represents.
- At least one specific design decision where the case study's informal
  description could have easily led to an un-tidy spreadsheet, and how
  your design avoids that trap.

### What to submit

Submit the following to Canvas:

1. Your spreadsheet file (.xlsx or a link to a Google Sheet with sharing
   enabled), containing only column headers filled in.
2. Your written justification (1 paragraph is sufficient).

## Week 5 Assignment: Querying the Bird Nest Box Database

This assignment uses a new practice database, `birds.sqlite`, covering a
nest box study with three habitat types (farmland, residential, forest).
Adults are banded, GPS-tagged, and have blood hormone samples and gut
microbiome samples taken. Hatchlings are banded and sexed at hatch,
measured regularly until fledging, and GPS-tagged just before fledging.

Unlike the in-class activity in Chapter 6, this time you're given five
complete, working queries. Your job isn't to write SQL from scratch, it's
to run each query, understand what it's doing, and interpret what the
result actually tells you biologically. This is just as important a skill
as writing queries in the first place: in practice, you'll often be
reading and adapting someone else's SQL (a labmate's, a collaborator's,
your own from six months ago) far more often than starting from a blank
page.

### Getting the data

`birds.sqlite` lives in the `data/` folder of the course repository,
alongside `frogs.sqlite`. If you already cloned the repo back in Chapter
6, just make sure you're up to date:

### Setup

```{r, eval=FALSE}
library(DBI)

con <- dbConnect(RSQLite::SQLite(), "data/birds.sqlite")
dbListTables(con)
```

### Query 1: Do males and females differ in nest attendance during incubation?

Neither `nests` nor `gps_fixes` has a column that says "this bird was
incubating right now." Instead, we infer it: a GPS fix very close to the
nest box's coordinates (within a small distance threshold) during the
incubation period (before the eggs hatch) is treated as "at the nest."
Everything else counts as "away." For the purposes of this example, the 
researchers set a distance threshold of 0.003 degrees from the nest box to 
determine at the nest vs. away.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    b.sex,
    SUM(CASE WHEN (ABS(g.lat - nb.latitude) + ABS(g.lon - nb.longitude)) < 0.003
             THEN 1 ELSE 0 END) AS nest_fixes,
    COUNT(*) AS total_fixes,
    ROUND(100.0 * SUM(CASE WHEN (ABS(g.lat - nb.latitude) + ABS(g.lon - nb.longitude)) < 0.003
                           THEN 1 ELSE 0 END) / COUNT(*), 1) AS pct_at_nest
  FROM gps_fixes g
  JOIN deployments d ON g.tracker_id = d.tracker_id
  JOIN birds b ON d.bird_id = b.bird_id
  JOIN nests n ON b.bird_id IN (n.female_id, n.male_id)
  JOIN nest_boxes nb ON n.box_id = nb.box_id
  WHERE g.timestamp <= n.hatch_date
    AND b.entry_stage != 'hatchling'
  GROUP BY b.sex;
")
```

**Question:** Based on the output, do males or females spend more time at
the nest during incubation? Does this match what you'd expect
biologically, or does it surprise you? 

**Question:** The distance threshold, `0.003` degrees, is a simplification,
real studies would use a proper distance calculation and a threshold
chosen based on GPS accuracy and nest box spacing.

Write a query that finds the distance between every pair of nest boxes,
and identifies the two closest boxes and how far apart they are. (Hint:
this requires joining the `nest_boxes` table to itself, a **self-join**,
something we haven't explicitly covered yet. You can join a table to
itself just like any other join, just give each copy a different alias,
e.g. `FROM nest_boxes a JOIN nest_boxes b ON a.box_id < b.box_id`, then
compute a distance between `a`'s and `b`'s coordinates. The `<` in the
`ON` clause, rather than `!=`, avoids getting every pair twice and avoids
comparing a box to itself.)

<!--
ANSWER KEY (not visible in rendered output)

SELECT
  a.box_id AS box_a,
  b.box_id AS box_b,
  ROUND(ABS(a.latitude - b.latitude) + ABS(a.longitude - b.longitude), 4) AS distance
FROM nest_boxes a
JOIN nest_boxes b ON a.box_id < b.box_id
ORDER BY distance ASC
LIMIT 5;

Closest pair: boxes 1 and 2, distance 0.006 (only 2x the 0.003 threshold).
-->

Once you have that result, answer: is the `0.003` threshold safely
smaller than the distance between the closest two boxes, or is it cutting
it close? What would happen to a GPS fix that happened to fall roughly
halfway between those two closest boxes, and why might that be a problem
for this specific pair, even if it's not a problem for most other box
pairs in the dataset?

### Query 2: Are some GPS tracker models more prone to being lost?

Just like "at the nest," "lost" isn't a column anywhere in this database.
A tracker is inferred to be lost when its deployment has no
`retrieved_date`, meaning it was never brought back in.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    t.tracker_model,
    COUNT(*) AS n_deployments,
    SUM(CASE WHEN d.retrieved_date IS NULL THEN 1 ELSE 0 END) AS n_lost,
    ROUND(100.0 * SUM(CASE WHEN d.retrieved_date IS NULL THEN 1 ELSE 0 END)
          / COUNT(*), 1) AS pct_lost
  FROM deployments d
  JOIN trackers t ON d.tracker_id = t.tracker_id
  GROUP BY t.tracker_model
  ORDER BY pct_lost DESC;
")
```

**Question:** Which tracker model has the highest loss rate? Given this
result, what would you recommend to the research team about which
tracker model to use going forward, and what else would you want to know
before making that call (e.g., cost, weight, battery life)?

**Question:** This query has one specific detail that's easy to overlook
but important for getting a correct result: the `100.0` (written as
`100.0`, not just `100`) in the `ROUND(100.0 * SUM(...) / COUNT(*), 1)`
line.

Which single line in the query is this, and what job is it doing? (Hint:
think about how SQLite handles division when both numbers involved are
whole numbers, versus when one of them has a decimal point.)

Try it yourself: change `100.0` to plain `100` in your query, rerun it,
and compare the `pct_lost` column to your original result. What changes,
and why? Which tracker model's row shows the clearest difference between
the two versions, and what does that tell you about when this kind of
mistake would be most likely to go unnoticed?

### Query 3: Does hatchling growth differ by habitat type?

This compares hatchling mass around day 15 of life (just before
fledging) across the three habitat types nest boxes are placed in.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    nb.location_type,
    ROUND(AVG(m.mass_g), 1) AS avg_mass_g,
    COUNT(*) AS n_records
  FROM morphometrics m
  JOIN birds b ON m.bird_id = b.bird_id
  JOIN nests n ON b.natal_nest_id = n.nest_id
  JOIN nest_boxes nb ON n.box_id = nb.box_id
  WHERE b.entry_stage = 'hatchling'
    AND julianday(m.measurement_date) - julianday(n.hatch_date) BETWEEN 13 AND 17
  GROUP BY nb.location_type
  ORDER BY avg_mass_g DESC;
")
```

**Question:** How much difference is there between habitat types? Given
the `n_records` column, do you think this dataset has enough
observations to draw a confident conclusion either way?

**Question:** This query needs to figure out how old each hatchling was,
in days, at the time of each measurement, using two date columns:
`m.measurement_date` and `n.hatch_date`. Which line does this, and what
does the `julianday()` function do?

Try running just this piece on its own to see what it returns:

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT measurement_date, julianday(measurement_date)
  FROM morphometrics
  LIMIT 5;
")
```

Dates in this database are stored as plain text (e.g., `'2023-05-03'`),
not as a special date type. Why can't you just subtract two date strings
directly, the way you might subtract two numbers? What would
`'2023-05-20' - '2023-05-03'` even mean to SQLite, and why does wrapping
each one in `julianday()` first solve that problem?

One more thing to notice: the query also has a `WHERE b.entry_stage =
'hatchling'` line. Given that the `JOIN` to `nests` uses
`b.natal_nest_id = n.nest_id`, and only hatchlings have a non-`NULL`
`natal_nest_id` (adults don't), is that `WHERE` line actually doing
anything here, or is it redundant? Explain your reasoning.

### Query 4: Does parent stress hormone level differ by habitat type?

Corticosterone is the primary stress hormone in birds, and the researchers 
hypothesize that bird parents in urban environments would show higher stress
levels than bird parents in forests.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    nb.location_type,
    ROUND(AVG(h.corticosterone), 2) AS avg_cort,
    COUNT(*) AS n_samples
  FROM hormones h
  JOIN birds b ON h.bird_id = b.bird_id
  JOIN nests n ON b.bird_id IN (n.female_id, n.male_id)
  JOIN nest_boxes nb ON n.box_id = nb.box_id
  GROUP BY nb.location_type
  ORDER BY avg_cort DESC;
")
```

**Question:** Which habitat type shows the highest average
corticosterone? Propose one biological explanation for why that habitat
might be more stressful for a nesting bird. Given the sample sizes shown,
how confident are you in this pattern?

**Question:** Look at the `JOIN` line connecting `birds` to `nests`:

```
JOIN nests n ON b.bird_id IN (n.female_id, n.male_id)
```

This looks a little different from the joins we've used so far, it uses `IN` instead of a plain `=`. Why can't this line just say `ON b.bird_id = n.female_id`? What would that simpler version miss, and why does `IN (n.female_id, n.male_id)` fix it?

### Query 5: Does gut microbiome richness differ between adults and hatchlings?

This counts the number of distinct OTUs (a proxy for microbial diversity)
detected per bird, then compares the average across adults and
hatchlings.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    b.entry_stage,
    ROUND(AVG(otu_count), 1) AS avg_otus_per_bird
  FROM (
    SELECT bird_id, COUNT(DISTINCT otu_id) AS otu_count
    FROM otu_reads
    GROUP BY bird_id
  ) sub
  JOIN birds b ON sub.bird_id = b.bird_id
  GROUP BY b.entry_stage;
")
```

**Question:** This query uses a subquery, a query nested inside another
query (the inner `SELECT` runs first, and the outer query treats its
result like a table). In your own words, explain what the inner subquery
computes, and why it needs to run before the outer `GROUP BY` can work.
Then interpret the result: do adults or hatchlings show higher microbial
richness, and can you think of a biological reason that might be true?

**Question:** Look at the inner subquery:

```
SELECT bird_id, COUNT(DISTINCT otu_id) AS otu_count
FROM otu_reads
GROUP BY bird_id
```

Try removing `GROUP BY bird_id` from just this inner piece, and run the whole query again. What happens to the result? Why does dropping this one line break the entire comparison between adults and hatchlings, even though the outer query still has its own `GROUP BY b.entry_stage` line? What is `GROUP BY bird_id` actually accomplishing here that the outer `GROUP BY` can't do on its own?

### What to submit

Submit an RMarkdown document (or plain R script) with all five queries,
their output, and your written answer to each question.
