# Assignments

This chapter collects each week's assignment in one place for easy
reference. New assignments will be added here as the semester progresses.

## Week 1: Software Setup and Final Project Topic Selection

Before our first class session, please install R and RStudio and verify
your installation.

## Software Requirements and Installation Instructions {-}

Before our first class session, please install the following software.
Come to class with everything installed and verified — we'll do a quick
check together, but coming prepared means we can dive into the material
right away.

### R {-}

R is a free software environment for statistical computing and graphics.
Installing or updating R is a separate process from installing or updating
RStudio (below) — you need both.

- **Windows:** Download the latest version at <https://cran.r-project.org/bin/windows/base/>
- **Mac OS:** Download the latest version at <https://cran.r-project.org/bin/macosx/>
- **Linux:** Follow the instructions for your distribution at <https://cloud.r-project.org/bin/linux/>

### RStudio {-}

RStudio is a free Integrated Development Environment (IDE) for R. Installing
or updating RStudio is separate from installing or updating R.

Download it from <https://posit.co/download/rstudio-desktop/>.

## The RStudio Interface {-}

When you open RStudio, you'll see the window divided into four panes. Each
one serves a different purpose:

**Source (top left)**
This is where you write and edit scripts, RMarkdown documents, and other
files. Code you write here isn't run automatically — you need to send it to
the Console (below) to execute it. Keeping your code in the Source pane,
rather than typing directly into the Console, is what makes your work
reproducible: it's saved, shareable, and can be re-run from start to finish.

**Console (bottom left)**
This is where R actually executes code and shows you the results. You can
type commands directly into the Console, but anything you type there isn't
saved anywhere — so it's best used for quick, one-off checks rather than your
actual analysis.

**Environment / History (top right)**
The Environment tab shows you all the objects (data frames, variables,
functions, etc.) currently loaded in your R session, along with their type
and a preview of their contents. The History tab keeps a running log of every
command you've executed in the Console during your session.

**Files / Plots / Packages / Help (bottom right)**
This pane is multi-purpose. *Files* lets you browse your project's folder
structure. *Plots* displays any graphs or visualizations you create. *Packages*
shows which R packages are installed and lets you load or update them. *Help*
displays documentation when you look up a function.

You can resize any pane by dragging its borders, and rearrange their
positions under **Tools > Global Options > Pane Layout** if you'd prefer a
different arrangement.

### Required R Packages {-}

We'll be using the following R packages throughout the course: `tidyverse`,
`rmarkdown`, `bookdown`, and `renv`. Install them, along with their dependencies, by running the
following in the R console:

```r
install.packages(c("tidyverse",
                    "rmarkdown",
                    "bookdown",
                    "renv"), dependencies = TRUE)
```

### Verifying Your Installation {-}

To confirm everything is working, open RStudio and run the following in
the console:

```r
R.version.string
library(tidyverse)
library(bookdown)
library(renv)
```

If R prints a version number and the `tidyverse` loads without error
messages, you're all set.

### Dataset eligibility criteria

Your dataset must meet the following criteria:

- **Public availability:** The dataset must be publicly accessible — for
  example, from a government agency, public data repository, or a published
  study's supplementary materials. No proprietary, restricted-access, or
  personally identifiable data.
- **Minimum size:** At least **100 rows (observations)** and **10 variables
  (columns)**.
- **Rationale:** A short written explanation (3–5 sentences) of why this
  dataset interests you and what you hope to learn from it.
- **Exploratory questions:** At least 2–3 questions you want to answer using
  this dataset by the end of the course.


### What to submit

Go to Canvas, and submit a screen shot of your R console output for the 
"Week 1" assignment that shows your successful download.

Then in the text box, submit the following:

- A brief description of your dataset, including an overview of how the data were collected, how many observations there are, and how many variables were measured.
- A short written explanation (3–5 sentences) of why this dataset interests you and what you hope to learn from it.
- At least 2–3 questions you want to answer using this dataset by the end of the course.
- A brief description on any data cleaning you anticipate needing to do, based on an initial look at the file.

## Week 2: Project Directory Structure and Git Setup

Now that you've chosen your final project dataset, it's time to set up the
project itself as a properly structured, version-controlled repository.

### What to do

**1. Design your directory structure.**

Based on what we covered in Chapter 1, plan out a folder structure for your
final project. At minimum, your structure should separate raw data,
scripts, and output — but feel free to adapt it to fit your specific
project.

**2. Create the folders locally.**

Make a new folder for your project (name it something descriptive, e.g.
`final-project`), then create your subfolders inside it. You can do this in
Finder/File Explorer, or from the terminal. For example, from inside your
project folder:

```
mkdir data
mkdir data/raw
mkdir data/processed
mkdir scripts
mkdir output
mkdir output/figures
```

(Adjust folder names to match your own planned structure — this is just an
example.)

**3. Turn it into a Git repository.**

Navigate into your project folder in the terminal (recall the drag-and-drop
or Option+right-click tricks from Chapter 2 if you need to find the path),
then run:

```
git init
```

**4. Add a `.gitignore` file.**

Create a new plain-text file named exactly `.gitignore` in your project's
top-level folder (see Chapter 2 for instructions specific to your
operating system). At minimum, include the following:

```
.Rhistory
.Rproj.user/
.DS_Store
```

Add any additional rules relevant to your own project — for example, if
you'll be working with large raw data files you don't want tracked.

**5. Add a README.**

Create a new plain-text file named `README.md` in your project's top-level
folder. At minimum, it should include:

- Your project title and a one-paragraph description of what it's about
- A brief explanation of your folder structure (what lives where)
- A note on your data source (a link or citation)

**6. Make your first commit.**

From inside your project folder, run:

```
git add .
git commit -m "Initial project structure"
```

### What to submit

Submit a short document (PDF or Word) containing:

1. A screenshot showing your folder structure.
2. A brief paragraph (3–5 sentences) explaining your folder structure —
   what each top-level folder is for and why you organized it this way.

## Week 3: Creating a GitHub Repository for Your Project

In Week 2, you created a local Git repository for your final project. This
week, you'll put that repository on GitHub so it's backed up remotely and
ready for collaboration.

### What to do

**1. Create a new repository on GitHub.**

- Go to <https://github.com> and log in.
- Click the **+** icon in the top-right corner and select **New repository**.
- Fill out the repository details:
  - **Repository name:** Use a short, descriptive name for your project
    (e.g., `my-lastname-final-project`)
  - **Description:** One sentence describing your project topic
  - **Visibility:** Choose **Public** (unless your data requires restricted
    access, in which case choose **Private**)
  - **Do NOT** check "Add a README file," "Add .gitignore," or "Choose a
    license" — your local repo already has these files, and we'll connect
    it directly
- Click **Create repository**. GitHub will show you a page with setup
  instructions — keep this page open.

**2. Connect your local repo to GitHub.**

Copy the repository URL from the GitHub page (it will look like
`https://github.com/yourusername/your-repo-name.git`). Then, from inside
your project folder, run the following, replacing `<repo-url>` with the URL
you copied:

```
git remote add origin <repo-url>
git branch -M main
git push -u origin main
```

If prompted for a username and password, enter your GitHub username. For
the password, use a Personal Access Token (PAT) instead of your account
password (see Chapter 3 for a refresher on generating one).

**3. Verify it worked.**

Refresh your repository's page on GitHub and confirm that your project
files — including your `.gitignore` and `README.md` — appear there.

### What to submit

Submit the following:

1. A screenshot of your repository's GitHub page showing your files.
2. The URL to your GitHub repository.

## Week 4: Designing a Tidy Spreadsheet for a Bear Research Case Study

This week, you'll apply the tidy data principles from Chapter 4 to a new
scenario — not your own final project dataset, but a case study describing
a black bear research project. Real field data collection is often
described informally, in field notes or verbal summaries, rather than
handed to you as a ready-made spreadsheet. Your job is to translate a
messy, informal description of a research project into a properly
structured, tidy spreadsheet design.

### The case study

A research team is studying a black bear population in a mountainous study
area. Here's how they describe their data collection process:

> "We trap bears using culvert traps and snares placed at a handful of
> fixed trap sites scattered across the study area. Every time we catch a
> bear — whether it's the first time or a recapture — we record the date,
> which trap site it was caught at, and the bear's ID (new bears get a new
> ID; if we recognize an ear tag from a previous capture, we reuse that
> ID). At first capture, we record the bear's sex. Every time we handle a
> bear, whether it's a new capture or a recapture, we take its weight and
> body length, and we note its age class (cub, yearling, or adult). Over
> the course of the study, we catch dozens of different bears, some of
> them multiple times, at several different trap sites, across several
> field seasons."

### What to do

**1. Identify the variables and observations.**

Read through the case study carefully. For each piece of information being
collected, ask: what is being measured (a variable)? Then think about what
a single row in your table should represent — one bear? One capture event?
Be careful here: a bear that's caught three times shouldn't be squeezed
into a single row, since its weight, length, age class, trap site, and
date are all different at each capture.

**2. Design your table.**

Sketch out the column headers your table needs, keeping in mind that the
same bear ID will appear on more than one row (once per capture), while
other information (like sex) stays the same across all of that bear's
rows. This is expected and fine — it's how a single tidy table
accommodates repeated measurements on the same individual.

**3. Build the actual spreadsheet template.**

In Excel or Google Sheets, create your table with only column headers
filled in — you don't need to invent or enter any actual data rows. Column
headers should be computer-readable (following the naming conventions from
Chapter 1: no spaces, no special characters, consistent capitalization).

Pay close attention to the tidy data pitfalls from Chapter 4 as you design
your headers: don't cram two variables into one column (like age class and
sex), don't let a single cell need to hold more than one value, and make
sure your table's unit of observation is consistent row to row.

**4. Write a short justification.**

In a separate document (or as a second tab/sheet), briefly explain:

- What a single row in your table represents.
- At least one specific design decision where the case study's informal
  description could have easily led to an un-tidy spreadsheet, and how
  your design avoids that trap.

### What to submit

Submit the following to Canvas:

1. Your spreadsheet file (.xlsx or a link to a Google Sheet with sharing
   enabled), containing only column headers filled in.
2. Your written justification (1 paragraph is sufficient).

## Week 5 Assignment: Querying the Bird Nest Box Database

This assignment uses a new practice database, `birds.sqlite`, covering a
nest box study with three habitat types (farmland, residential, forest).
Adults are banded, GPS-tagged, and have blood hormone samples and gut
microbiome samples taken. Hatchlings are banded and sexed at hatch,
measured regularly until fledging, and GPS-tagged just before fledging.

Unlike the in-class activity in Chapter 6, this time you're given five
complete, working queries. Your job isn't to write SQL from scratch, it's
to run each query, understand what it's doing, and interpret what the
result actually tells you biologically. This is just as important a skill
as writing queries in the first place: in practice, you'll often be
reading and adapting someone else's SQL (a labmate's, a collaborator's,
your own from six months ago) far more often than starting from a blank
page.

### Getting the data

`birds.sqlite` lives in the `data/` folder of the course repository,
alongside `frogs.sqlite`. If you already cloned the repo back in Chapter
6, just make sure you're up to date:

### Setup

```{r, eval=FALSE}
library(DBI)

con <- dbConnect(RSQLite::SQLite(), "data/birds.sqlite")
dbListTables(con)
```

### Query 1: Do males and females differ in nest attendance during incubation?

Neither `nests` nor `gps_fixes` has a column that says "this bird was
incubating right now." Instead, we infer it: a GPS fix very close to the
nest box's coordinates (within a small distance threshold) during the
incubation period (before the eggs hatch) is treated as "at the nest."
Everything else counts as "away." For the purposes of this example, the 
researchers set a distance threshold of 0.003 degrees from the nest box to 
determine at the nest vs. away.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    b.sex,
    SUM(CASE WHEN (ABS(g.lat - nb.latitude) + ABS(g.lon - nb.longitude)) < 0.003
             THEN 1 ELSE 0 END) AS nest_fixes,
    COUNT(*) AS total_fixes,
    ROUND(100.0 * SUM(CASE WHEN (ABS(g.lat - nb.latitude) + ABS(g.lon - nb.longitude)) < 0.003
                           THEN 1 ELSE 0 END) / COUNT(*), 1) AS pct_at_nest
  FROM gps_fixes g
  JOIN deployments d ON g.tracker_id = d.tracker_id
  JOIN birds b ON d.bird_id = b.bird_id
  JOIN nests n ON b.bird_id IN (n.female_id, n.male_id)
  JOIN nest_boxes nb ON n.box_id = nb.box_id
  WHERE g.timestamp <= n.hatch_date
    AND b.entry_stage != 'hatchling'
  GROUP BY b.sex;
")
```

**Question:** Based on the output, do males or females spend more time at
the nest during incubation? Does this match what you'd expect
biologically, or does it surprise you? 

**Question:** The distance threshold, `0.003` degrees, is a simplification,
real studies would use a proper distance calculation and a threshold
chosen based on GPS accuracy and nest box spacing.

Write a query that finds the distance between every pair of nest boxes,
and identifies the two closest boxes and how far apart they are. (Hint:
this requires joining the `nest_boxes` table to itself, a **self-join**,
something we haven't explicitly covered yet. You can join a table to
itself just like any other join, just give each copy a different alias,
e.g. `FROM nest_boxes a JOIN nest_boxes b ON a.box_id < b.box_id`, then
compute a distance between `a`'s and `b`'s coordinates. The `<` in the
`ON` clause, rather than `!=`, avoids getting every pair twice and avoids
comparing a box to itself.)

<!--
ANSWER KEY (not visible in rendered output)

SELECT
  a.box_id AS box_a,
  b.box_id AS box_b,
  ROUND(ABS(a.latitude - b.latitude) + ABS(a.longitude - b.longitude), 4) AS distance
FROM nest_boxes a
JOIN nest_boxes b ON a.box_id < b.box_id
ORDER BY distance ASC
LIMIT 5;

Closest pair: boxes 1 and 2, distance 0.006 (only 2x the 0.003 threshold).
-->

Once you have that result, answer: is the `0.003` threshold safely
smaller than the distance between the closest two boxes, or is it cutting
it close? What would happen to a GPS fix that happened to fall roughly
halfway between those two closest boxes, and why might that be a problem
for this specific pair, even if it's not a problem for most other box
pairs in the dataset?

### Query 2: Are some GPS tracker models more prone to being lost?

Just like "at the nest," "lost" isn't a column anywhere in this database.
A tracker is inferred to be lost when its deployment has no
`retrieved_date`, meaning it was never brought back in.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    t.tracker_model,
    COUNT(*) AS n_deployments,
    SUM(CASE WHEN d.retrieved_date IS NULL THEN 1 ELSE 0 END) AS n_lost,
    ROUND(100.0 * SUM(CASE WHEN d.retrieved_date IS NULL THEN 1 ELSE 0 END)
          / COUNT(*), 1) AS pct_lost
  FROM deployments d
  JOIN trackers t ON d.tracker_id = t.tracker_id
  GROUP BY t.tracker_model
  ORDER BY pct_lost DESC;
")
```

**Question:** Which tracker model has the highest loss rate? Given this
result, what would you recommend to the research team about which
tracker model to use going forward, and what else would you want to know
before making that call (e.g., cost, weight, battery life)?

**Question:** This query has one specific detail that's easy to overlook
but important for getting a correct result: the `100.0` (written as
`100.0`, not just `100`) in the `ROUND(100.0 * SUM(...) / COUNT(*), 1)`
line.

Which single line in the query is this, and what job is it doing? (Hint:
think about how SQLite handles division when both numbers involved are
whole numbers, versus when one of them has a decimal point.)

Try it yourself: change `100.0` to plain `100` in your query, rerun it,
and compare the `pct_lost` column to your original result. What changes,
and why? Which tracker model's row shows the clearest difference between
the two versions, and what does that tell you about when this kind of
mistake would be most likely to go unnoticed?

### Query 3: Does hatchling growth differ by habitat type?

This compares hatchling mass around day 15 of life (just before
fledging) across the three habitat types nest boxes are placed in.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    nb.location_type,
    ROUND(AVG(m.mass_g), 1) AS avg_mass_g,
    COUNT(*) AS n_records
  FROM morphometrics m
  JOIN birds b ON m.bird_id = b.bird_id
  JOIN nests n ON b.natal_nest_id = n.nest_id
  JOIN nest_boxes nb ON n.box_id = nb.box_id
  WHERE b.entry_stage = 'hatchling'
    AND julianday(m.measurement_date) - julianday(n.hatch_date) BETWEEN 13 AND 17
  GROUP BY nb.location_type
  ORDER BY avg_mass_g DESC;
")
```

**Question:** How much difference is there between habitat types? Given
the `n_records` column, do you think this dataset has enough
observations to draw a confident conclusion either way?

**Question:** This query needs to figure out how old each hatchling was,
in days, at the time of each measurement, using two date columns:
`m.measurement_date` and `n.hatch_date`. Which line does this, and what
does the `julianday()` function do?

Try running just this piece on its own to see what it returns:

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT measurement_date, julianday(measurement_date)
  FROM morphometrics
  LIMIT 5;
")
```

Dates in this database are stored as plain text (e.g., `'2023-05-03'`),
not as a special date type. Why can't you just subtract two date strings
directly, the way you might subtract two numbers? What would
`'2023-05-20' - '2023-05-03'` even mean to SQLite, and why does wrapping
each one in `julianday()` first solve that problem?

One more thing to notice: the query also has a `WHERE b.entry_stage =
'hatchling'` line. Given that the `JOIN` to `nests` uses
`b.natal_nest_id = n.nest_id`, and only hatchlings have a non-`NULL`
`natal_nest_id` (adults don't), is that `WHERE` line actually doing
anything here, or is it redundant? Explain your reasoning.

### Query 4: Does parent stress hormone level differ by habitat type?

Corticosterone is the primary stress hormone in birds, and the researchers 
hypothesize that bird parents in urban environments would show higher stress
levels than bird parents in forests.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    nb.location_type,
    ROUND(AVG(h.corticosterone), 2) AS avg_cort,
    COUNT(*) AS n_samples
  FROM hormones h
  JOIN birds b ON h.bird_id = b.bird_id
  JOIN nests n ON b.bird_id IN (n.female_id, n.male_id)
  JOIN nest_boxes nb ON n.box_id = nb.box_id
  GROUP BY nb.location_type
  ORDER BY avg_cort DESC;
")
```

**Question:** Which habitat type shows the highest average
corticosterone? Propose one biological explanation for why that habitat
might be more stressful for a nesting bird. Given the sample sizes shown,
how confident are you in this pattern?

**Question:** Look at the `JOIN` line connecting `birds` to `nests`:

```
JOIN nests n ON b.bird_id IN (n.female_id, n.male_id)
```

This looks a little different from the joins we've used so far, it uses `IN` instead of a plain `=`. Why can't this line just say `ON b.bird_id = n.female_id`? What would that simpler version miss, and why does `IN (n.female_id, n.male_id)` fix it?

### Query 5: Does gut microbiome richness differ between adults and hatchlings?

This counts the number of distinct OTUs (a proxy for microbial diversity)
detected per bird, then compares the average across adults and
hatchlings.

```{r, eval=FALSE}
dbGetQuery(con, "
  SELECT
    b.entry_stage,
    ROUND(AVG(otu_count), 1) AS avg_otus_per_bird
  FROM (
    SELECT bird_id, COUNT(DISTINCT otu_id) AS otu_count
    FROM otu_reads
    GROUP BY bird_id
  ) sub
  JOIN birds b ON sub.bird_id = b.bird_id
  GROUP BY b.entry_stage;
")
```

**Question:** This query uses a subquery, a query nested inside another
query (the inner `SELECT` runs first, and the outer query treats its
result like a table). In your own words, explain what the inner subquery
computes, and why it needs to run before the outer `GROUP BY` can work.
Then interpret the result: do adults or hatchlings show higher microbial
richness, and can you think of a biological reason that might be true?

**Question:** Look at the inner subquery:

```
SELECT bird_id, COUNT(DISTINCT otu_id) AS otu_count
FROM otu_reads
GROUP BY bird_id
```

Try removing `GROUP BY bird_id` from just this inner piece, and run the whole query again. What happens to the result? Why does dropping this one line break the entire comparison between adults and hatchlings, even though the outer query still has its own `GROUP BY b.entry_stage` line? What is `GROUP BY bird_id` actually accomplishing here that the outer `GROUP BY` can't do on its own?

### What to submit

Submit an RMarkdown document (or plain R script) with all five queries,
their output, and your written answer to each question.

## Week 6: Design Your Final Project Database Structure

Now that you've worked through relational databases (Chapter 5), SQL
(Chapter 6), and building a database from CSVs in R (Chapter 7), it's
time to apply all of that to your own final project. This week, you'll
design the table structure your project's data would actually need,
without building the database itself yet.

### What to do

**1. Revisit your dataset.**

Go back to your Week 1 proposal: your dataset, your rationale, and your
exploratory questions. Keep these in mind as you design, since your
table structure should support the questions you actually want to ask.

**2. Decide how many tables you need.**

Using the principles from Chapter 5, work through whether your data
belongs in one table or several:

- What is the sampling unit (or units) in your data? A single
  consistent unit usually means a single table is enough. Multiple
  distinct units (e.g., a site, and repeated visits to that site) often
  means multiple tables.
- Is there a mix of static information (recorded once) and dynamic
  information (recorded repeatedly)? If so, consider splitting them,
  the same logic behind separating a frog's sex from its repeated
  weight measurements.
- Does every column hold one atomic value? If you find yourself wanting
  to cram a list into a single cell, that's usually a sign you need an
  additional table (a junction table, if the relationship is many to
  many, as in our diet_asvs / diet_asv_reads example).

A single table is a completely acceptable answer here, if that's what
your data actually calls for. What matters is that you can justify the
decision using these principles, not that you force complexity your
project doesn't need.

**3. Map out each table.**

For every table your design requires, write out:

- The table name
- Every column header, with its data type (text, integer, float, date,
  boolean)
- Which column is the primary key
- Any foreign keys, and which table and column each one references

**4. If you have more than one table, explain the relationships.**

For each pair of related tables, state whether the relationship is one
to many or many to many, and briefly explain why. If it's many to many,
identify the junction table that handles it.

**5. Draw a diagram.**

Using PowerPoint, Google Slides, or a similar tool, draw a diagram of
your table structure, following the same style as the diagrams in
Chapters 5 through 7:

- One box (or text box) per table, with the table name at the top.
- List every column inside its box, and mark the primary key (e.g.,
  underline it, or add "(PK)" next to it) and any foreign keys (add
  "(FK)" next to them).
- Draw a line or arrow connecting each pair of related tables, from the
  primary key in one table to the matching foreign key in the other.
- Label each connecting line as "1 to many" or "many to many," matching
  the relationships you identified in step 4.

If you have a single table, you don't need any connecting lines, just
one box showing its columns.

### What to submit

Submit a short document (PDF, Word, or a rendered RMarkdown file)
containing:

1. Your diagram (a PowerPoint or Google Slides screenshot/export is
   fine, it doesn't need to be a separate file).
2. Your table map in writing: every table, its columns and data types,
   its primary key, and any foreign keys (this can be redundant with
   your diagram, that's fine, the written version should be
   detailed enough to actually build the database from later).
3. A short paragraph (3-5 sentences) explaining your reasoning: why this
   many tables, and how they relate to each other (or, if you landed on
   a single table, why that's the right call for your data).

## Week 7: Build Your Final Project Database

Last week, you designed the tables your final project needs. This week, you'll build them. As Chapter 5 explained, a database is a set of related tables, and you don't need SQL to have one. You'll write one R script that reads your raw data, splits it into the tables in your design, saves each table as a CSV file, and checks the result. Because the whole process lives in a script, anyone can rebuild your tables from your raw data by running it again, including you, a month from now. That is what makes them reproducible.

There are two tracks. **Track A** saves your tables as CSV files and checks them with R, and it's what most of you should do. **Track B** is optional: it builds a SQLite database instead, as in Chapters 6 and 7, which is worth choosing if you want to query your data with SQL, or if your design has several linked tables and you want the database to enforce the links. Both tracks start the same way and check the same things. Everything in the numbered steps below is Track A, and Track B has its own section after them.

### Before you start

You'll need:

- Your Week 6 table map and diagram.
- Your final project repository from Weeks 2 and 3, on your computer, with your raw data file or files saved in `data/raw/` and left unedited (Chapter 1: never edit raw data by hand).
- For Track B only, the `DBI` and `RSQLite` packages, which you installed in Chapter 6.

This assignment expects your raw data to be CSV files, which is what most public data portals provide. If yours are Excel files, save a CSV copy of each sheet you need into `data/raw/` and keep the originals too.

You'll write your code in a script file and run it from the **Console**, the pane in RStudio where R prints its results, marked by a `>` prompt. Paste commands into the Console or into your script, never into an `.Rmd` file.

### What to do

**1. Run the worked example.**

Before you touch your own data, see what a successful build looks like. In RStudio, create a new script (**File > New File > R Script**), paste in the following, and click **Source** at the top of the script. It uses a small dataset that comes with R (the carbon dioxide uptake of 12 plants, measured 7 times each) and splits it into two related tables, `plants` and `measurements`, just as you might split a frog's fixed information from its repeated measurements (Chapter 5). It saves its files in a temporary folder, so nothing is saved on your computer.

```{r, eval=FALSE}
# 1. Read the raw data ---------------------------------------------------------
# This example uses a dataset that comes with R, so it runs as written.
# In your own script, you will read your own file or files instead (see step 3).
co2 <- as.data.frame(CO2)
co2$Plant <- as.character(co2$Plant)
co2$Type <- as.character(co2$Type)
co2$Treatment <- as.character(co2$Treatment)

# 2. Split the raw data into tables --------------------------------------------
# Table 1: plants. One row per plant, holding information that never changes.
plants <- unique(co2[, c("Plant", "Type", "Treatment")])
names(plants) <- c("plant_id", "origin", "treatment")

# Table 2: measurements. One row per measurement, with a unique identifier (measurement_id) as its key.
measurements <- co2[, c("Plant", "conc", "uptake")]
names(measurements) <- c("plant_id", "co2_concentration", "co2_uptake")
measurements$measurement_id <- 1:nrow(measurements)
measurements <- measurements[, c("measurement_id", "plant_id", "co2_concentration", "co2_uptake")]

# 3. Save each table as a CSV file ---------------------------------------------
# This example saves its files in a temporary folder. In your own script, use
# "data/processed" instead, the folder you made in Week 2.
out_dir <- file.path(tempdir(), "tables")
dir.create(out_dir, showWarnings = FALSE)
write.csv(plants, file.path(out_dir, "plants.csv"), row.names = FALSE, na = "")
write.csv(measurements, file.path(out_dir, "measurements.csv"), row.names = FALSE, na = "")

# 4. Check the tables ----------------------------------------------------------
# Read the files back from disk, so that you are checking what was actually saved
plants <- read.csv(file.path(out_dir, "plants.csv"), stringsAsFactors = FALSE, na.strings = "")
measurements <- read.csv(file.path(out_dir, "measurements.csv"), stringsAsFactors = FALSE, na.strings = "")

# The number of rows in each table
cat("plants:", nrow(plants), "rows\n")
cat("measurements:", nrow(measurements), "rows\n")

# Keys that appear more than once (each of these should be 0)
cat("Duplicated plant_id values in plants:", sum(duplicated(plants$plant_id)), "\n")
cat("Duplicated measurement_id values in measurements:", sum(duplicated(measurements$measurement_id)), "\n")

# Rows that point to something that doesn't exist (this should be 0)
cat("Measurements whose plant_id is not in plants:", sum(!(measurements$plant_id %in% plants$plant_id)), "\n")

# The number of missing values in each column
print(colSums(is.na(plants)))
print(colSums(is.na(measurements)))

# A summary that uses both tables, to prove the relationship works
combined <- merge(measurements, plants, by = "plant_id")
print(aggregate(co2_uptake ~ treatment, data = combined, FUN = mean))
```

When it runs, you should see this in the Console:

```
plants: 12 rows
measurements: 84 rows
Duplicated plant_id values in plants: 0
Duplicated measurement_id values in measurements: 0
Measurements whose plant_id is not in plants: 0
 plant_id    origin treatment
        0         0         0
   measurement_id          plant_id co2_concentration        co2_uptake
                0                 0                 0                 0
   treatment co2_uptake
1    chilled   23.78333
2 nonchilled   30.64286
```

The script has four sections, and the next steps ask you to adapt each one. Section 4 is the important one. A CSV file will happily store a row that points to a plant that doesn't exist, or the same key twice, and it won't tell you. A database system such as SQLite can refuse those rows. So here, your checks do that job: they count the problems, and every count should be 0. Missing values can be legitimate, so those counts only need to make sense.

**2. Start your own script.**

Open a new script (**File > New File > R Script**), paste in the example above, and save it as `01-build-database.R` in the `scripts` folder of your final project repository. You'll edit this copy in the steps below. Leave the example in your first script untouched, so you can refer back to it.

**3. Read your raw data (section 1).**

Replace the example code in section 1 with code that reads your own raw data. What to write depends on how your data came.

*One file.* Read it into a data frame called `raw`, using your file's name:

```{r, eval=FALSE}
raw <- read.csv("data/raw/my-data.csv", stringsAsFactors = FALSE, na.strings = "")
```

*Several files, each of which is already its own table* (for example, one file of sites and another of observations). Read each one into its own data frame, named for what it holds:

```{r, eval=FALSE}
sites_raw <- read.csv("data/raw/sites.csv", stringsAsFactors = FALSE, na.strings = "")
observations_raw <- read.csv("data/raw/observations.csv", stringsAsFactors = FALSE, na.strings = "")
```

*Several files that are pieces of one table* (for example, one file per year, all with the same columns). Read them all and stack them into one data frame. The first line lists every file in `data/raw` whose name matches the pattern, so change the pattern to match only the files you want. The second line prints the list, so you can check it before going on:

```{r, eval=FALSE}
files <- list.files("data/raw", pattern = "^survey-.*\\.csv$", full.names = TRUE)
print(files)
raw <- do.call(rbind, lapply(files, read.csv, stringsAsFactors = FALSE, na.strings = ""))
```

If this stops with `names do not match previous names`, one of the files names its columns differently from the others. Read that file on its own and fix its column names with `names()` before stacking.

Sometimes a piece of information, such as the year, appears only in the file names. In that case, record each row's file name in a column as you read the files, and then pull the year out of it. This version works when the only digits in the file names are the year, as in `survey-2022.csv`:

```{r, eval=FALSE}
read_one <- function(file) {
  data <- read.csv(file, stringsAsFactors = FALSE, na.strings = "")
  data$source_file <- basename(file)
  data
}
raw <- do.call(rbind, lapply(files, read_one))
raw$year <- as.integer(gsub("[^0-9]", "", raw$source_file))
```

The `na.strings = ""` part tells R to treat an empty cell as a missing value (you met this in Chapter 7's activity). Every path starts with `data/raw/`, a relative path from the top of your project folder, as in Chapter 1.

Whichever case applies, run your code, then look at what you read in, so you know the starting point: `nrow()` tells you the number of rows of each data frame, and `str()` shows each column and the kind of data R found in it. Write down the number of rows of each data frame. You'll compare them to your tables later.

Look closely at the column names, using `names()`. When R reads a CSV, it replaces spaces and other special characters in column names with dots, so a column called `Water Temp C` becomes `Water.Temp.C`. Those dotted names are the ones to use when you split your data in the next step.

**4. Split your data into tables (section 2).**

Use your Week 6 table map. Each table in your design either already exists as its own data frame, because it came as its own file, or comes from splitting a data frame. Section 2 of the example shows the splitting pattern, and you'll repeat it for each table that needs it. (The examples call the data frame `raw`, so use the names of your own data frames.)

- A table with **one row per thing** (such as one row per plant, site, or individual) comes from `unique(raw[, c(...)])`, which keeps the columns you name and removes the repeated rows. If the table has only one column, write `drop = FALSE` after the column name, as in `unique(raw[, c("Site.Name"), drop = FALSE])`, so that R keeps it as a table.
- A table with **repeated rows** (such as measurements) comes from `raw[, c(...)]`, which keeps the columns you name and every row.
- `names(table) <- c(...)` gives the columns their final names. Use lowercase letters and underscores, with no spaces, as in Chapter 1.
- A table that has no natural key gets a serial number, like `measurement_id` in the example.

If your Week 6 design was a **single table**, skip the splitting. Give your data frame the final column names, and add a serial key if it has no column that uniquely identifies each row:

```{r, eval=FALSE}
raw$record_id <- 1:nrow(raw)
```

A table that already came as its own file doesn't need splitting. Give its columns their final names with `names()`, and add a serial key if it has no column that uniquely identifies each row, as above.

Dates need a little care. Write them as `YYYY-MM-DD`, which sorts correctly and is unambiguous, as in the databases in Chapters 5 through 7. If your dates look like `5/1/2024`, convert them (the letters after each `%` describe the format of your dates):

```{r, eval=FALSE}
raw$date <- as.character(as.Date(raw$date, format = "%m/%d/%Y"))
```

If your design needs a key that your raw data doesn't have, such as an ID for each visit to a site, create it in `raw` before you split, so that every table gets exactly the same value. For example, this builds an ID like `Alder Pond_2024-05-01` from a site column and a date column (use your own column names, and do it after converting your dates):

```{r, eval=FALSE}
raw$visit_id <- paste(raw$Site.Name, raw$Visit.Date, sep = "_")
```

**5. Save your tables (section 3).**

Replace the two `write.csv()` lines with one for each of your tables, and change `out_dir` to `"data/processed"`. The file name should be the table's name, as in `plants.csv`. Two settings matter. `row.names = FALSE` stops R from adding an unwanted column of row numbers. `na = ""` writes a missing value as an empty cell, which `read.csv(..., na.strings = "")` reads back as missing.

**6. Check your tables (section 4).**

Adapt the checks to your tables: read each saved file back, count its rows, and count the duplicated keys. If your design has more than one table, also count the rows in each child table whose key doesn't exist in its parent table. Then run `str()` on each table and confirm that every column has the type you expect: `chr` for text, `int` for whole numbers, and `num` for numbers with decimals. Replace the final summary with one of your own that uses your tables and moves you toward one of your exploratory questions from Week 1. With several tables, join at least two of them using `merge()`, as in the example. With one table, use `aggregate()` or `table()`.

The row counts should make sense. If each row of a data frame you read in became one row of a table, that table should have the same number of rows as that data frame, which you wrote down in step 3.

**7. Run your script from the top.**

Your script uses relative paths, which only work when R is working in the top folder of your project. Set that folder in the Console, then run the script. Replace the path with the path to your project folder (Chapters 2 and 6 show how to find it):

```{r, eval=FALSE}
setwd("/path/to/your/final-project")
source("scripts/01-build-database.R")
```

Don't put `setwd()` inside the script. A path that works on your computer wouldn't work on anyone else's, and the script has to work for anyone (Chapter 1). Once it runs without errors, run the last two lines again. The output should be identical, because the script overwrites its own files.

**8. Commit and push.**

Your processed CSV files are plain text, they're small, and they let anyone see your tables on GitHub without running anything, so commit them along with your script. (If one of your files is very large, tens of megabytes, add it to `.gitignore` and tell your instructor.) In a terminal inside your project folder, run:

```
git status
git add .
git commit -m "Add script that builds the project tables"
git push
```

`git status` should list `scripts/01-build-database.R` and your CSV files in `data/processed`. Then open your repository on GitHub and confirm that they're there.

### If something goes wrong

Most mistakes show up as a short R message. The ones you're most likely to see:

- **`cannot open file 'data/raw/...': No such file or directory`.** R isn't working in your project folder, or the file name is wrong. Run `getwd()` to check where R is, and `setwd()` to move it.
- **`cannot open file 'data/processed/...': No such file or directory`.** The `data/processed` folder doesn't exist, or R isn't working in your project folder. Make sure the `dir.create()` line comes before your `write.csv()` lines.
- **`names do not match previous names`.** You are stacking several files, and one of them names its columns differently from the others. See step 3.
- **`undefined columns selected`.** A column name inside `raw[, c(...)]` doesn't match any column in `raw`. Compare it with `names(raw)`, and remember that spaces became dots.
- **`object '...' not found`.** A name is misspelled, or you skipped the line that creates it. Run the script from the top.

Your checks print numbers rather than errors, so a problem shows up as a count that isn't 0:

- **Duplicated keys.** The table needs `unique()` applied to the right columns. If the same ID has different details on different rows, your raw data has an inconsistency to resolve first.
- **Rows whose key doesn't exist in the other table.** The two tables were built from different values, or an ID has a typo or an extra space. Compare the IDs in the two tables.

### Optional Track B: build a SQLite database

Choose this track if you'd like to query your tables with SQL, as in Chapter 6, or if you want the database itself to refuse rows that break a relationship between tables. Steps 1 through 4 are the same, so you read your data and split it into tables exactly as above. Then replace sections 3 through 5 of your script with the following, which builds the database with the tools from Chapter 7. To see it work first, add it after sections 1 and 2 of the worked example:

```{r, eval=FALSE}
library(DBI)

# If an earlier run of this script stopped with an error, close its connection first
if (exists("con")) try(dbDisconnect(con), silent = TRUE)

# 3. Create an empty database and its tables -----------------------------------
# This example builds its database in a temporary file. In your own script, use
# a permanent location instead, for example "data/processed/final-project.sqlite".
db_path <- tempfile(fileext = ".sqlite")
unlink(db_path)  # start from nothing, so the script can be run again and again
con <- dbConnect(RSQLite::SQLite(), db_path)
dbExecute(con, "PRAGMA foreign_keys = ON;")

# Create the table that others point to first (plants), then the table that points to it
dbExecute(con, "CREATE TABLE plants (
  plant_id TEXT PRIMARY KEY,
  origin TEXT NOT NULL,
  treatment TEXT NOT NULL
);")

dbExecute(con, "CREATE TABLE measurements (
  measurement_id INTEGER PRIMARY KEY,
  plant_id TEXT NOT NULL,
  co2_concentration REAL,
  co2_uptake REAL,
  FOREIGN KEY (plant_id) REFERENCES plants(plant_id)
);")

# 4. Load the data into the tables ---------------------------------------------
dbWriteTable(con, "plants", plants, append = TRUE)
dbWriteTable(con, "measurements", measurements, append = TRUE)

# 5. Check the database --------------------------------------------------------
print(dbListTables(con))

for (table in dbListTables(con)) {
  n <- dbGetQuery(con, paste0("SELECT COUNT(*) AS n FROM ", table, ";"))$n
  cat(table, ":", n, "rows\n")
}

# The number of rows that break a relationship between tables (this should be 0)
broken <- dbGetQuery(con, "PRAGMA foreign_key_check;")
cat("Rows that break a relationship:", nrow(broken), "\n")

# A query that uses both tables, to prove the relationship works
print(dbGetQuery(con, "
  SELECT plants.treatment, AVG(measurements.co2_uptake) AS mean_uptake
  FROM measurements
  JOIN plants ON measurements.plant_id = plants.plant_id
  GROUP BY plants.treatment;
"))

dbDisconnect(con)
```

It should print the table names, `measurements : 84 rows`, `plants : 12 rows`, `Rows that break a relationship: 0`, and a small table of mean uptake for each treatment.

Two lines are new compared with Chapter 7. `PRAGMA foreign_keys = ON` tells SQLite to actually enforce the relationships between your tables. It only does so if you ask, every time you connect, and without it a table could point to a row that doesn't exist. `unlink(db_path)` deletes any old copy of the database before building a new one, so you can run the script as many times as you like without a "table already exists" error.

To adapt it to your project:

- **Create your tables.** Replace the two `CREATE TABLE` statements with one for each table in your design. For each column, choose a data type from this table:

| Kind of data | Type to use |
|--------------|-------------|
| Text, categories, and IDs that contain letters | `TEXT` |
| Whole numbers, such as counts and years | `INTEGER` |
| Numbers with decimals, such as measurements | `REAL` |
| Dates and times | `TEXT`, written as `YYYY-MM-DD` or `YYYY-MM-DD HH:MM:SS` |
| TRUE and FALSE | `INTEGER`, using `1` and `0` |

- **Add your rules.** `PRIMARY KEY` goes on the column that uniquely identifies each row. `NOT NULL` goes on any column that must always have a value. A `FOREIGN KEY (column) REFERENCES other_table(column)` line goes in each table that points to another one. Create the tables that others point to first.
- **Choose a permanent location.** Change `db_path` to `"data/processed/final-project.sqlite"`. Keep the `unlink()`, `dbConnect()`, and `PRAGMA` lines as they are.
- **Load your tables.** Write one `dbWriteTable(con, "table_name", data_frame, append = TRUE)` line for each table, loading the tables that others point to first. The column names in each data frame must match the column names in its table. The order of the columns doesn't matter.
- **Check your database.** Keep the first three checks, and replace the final query with one of your own that uses your tables and moves you toward one of your exploratory questions from Week 1 (Chapter 6 shows how).
- **Commit and push.** A SQLite file is a build product, like the website in Chapter 9: your script recreates it from your raw data whenever it's needed, so it doesn't need its own tracked history (Chapter 2). Open your project's `.gitignore` file in RStudio (the Files pane lists it, or see Chapter 2 for showing hidden files), add a new line containing `*.sqlite`, and save it. Then commit and push your script as in step 8. `git status` should list the script and no `.sqlite` file.

Track B has its own error messages, because SQLite is strict about your rules:

- **`unable to open database file`.** The folder in `db_path` doesn't exist. Create `data/processed`, or check that R is working in your project folder.
- **`UNIQUE constraint failed: table.column`.** Two rows have the same value in a key column. Apply `unique()` to the right columns, as in Track A.
- **`FOREIGN KEY constraint failed`.** Either a table that points to another was loaded before it, or a row refers to an ID that doesn't exist in the other table. Load the other table first, and check for typos in the IDs.
- **`NOT NULL constraint failed: table.column`.** A column marked `NOT NULL` has missing values. Fix the data, or remove `NOT NULL` from that column if values may truly be missing.
- **`Columns ... not found`.** A column name in your data frame doesn't match any column in the table. Compare `names(your_data_frame)` with your `CREATE TABLE` statement.
- **`table ... already exists`.** An earlier run stopped partway and left its connection open. Run `dbDisconnect(con)`, then run the script again.

### What to submit

Submit the following:

1. The link to your final project repository on GitHub, with `scripts/01-build-database.R` pushed, and the track you used (A or B).
2. A document (PDF, Word, or a rendered RMarkdown file) containing the output your script prints, copied from the Console: the number of rows in each table, the check results, and the result of your own summary or query.
3. A short paragraph (3 to 5 sentences) describing how your tables differ from your Week 6 design, if they do, and what you changed and why. If they don't, describe one thing you had to figure out while building them.

<!--
INSTRUCTOR NOTES (not visible in rendered output)

To check a submission, clone the student's repository, set the working directory in R to the cloned folder with setwd, and run source("scripts/01-build-database.R"). It should finish without errors and print the number of rows in each table, the results of the integrity checks, and the result of the student's own summary or query. Run it a second time to confirm the output is the same. Track A leaves one CSV file per table in data/processed, which are committed. Track B builds a SQLite file in data/processed, which is ignored by Git, so it appears only after the script is run.

Things to look for in both tracks: the script uses relative paths and contains no setwd call; the raw data is unedited; every table has a key; the checks report 0 duplicated keys and 0 rows pointing to a missing parent row; the row counts make sense given the raw data; the final summary uses the student's own tables.
Track A: the CSV files are written with row.names = FALSE, dates are written as YYYY-MM-DD, and the checks read the saved files back rather than the objects in memory.
Track B: PRAGMA foreign_keys is turned on; parent tables are created and loaded before the tables that point to them; the script starts from nothing (unlink on the database path) so it can be rerun.

Single table designs are acceptable if justified in Week 6. They should still save one CSV file (or build one table) and run the same checks.
-->
